### 2013-10-23 src2/thesis.tex
update

\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Computation}
\newcommand{\dept}{Department of Computer Science and Engineering}
\newcommand{\uni}{Chalmers University of Technology and G\"oteborg University}
\newcommand{\group}{Functional Programming Research Group}

@@

\begin{enumerate}[A.] 
\item Josef Svenningsson and Bo Joel Svensson. Simple and Compositional Reification of Monadic Embedded Languages, 2013. \emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
\item Bo Joel Svensson, Mary Sheeran and Koen Claessen. Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors. In \emph{Proc. of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
\item Bo Joel Svensson, Koen Claessen and Mary Sheeran. GPGPU kernel implementation and refinement using Obisidan. \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
\item Koen Claessen, Mary Sheeran and Bo Joel Svensson. Expressive Array Constructs in an Embedded GPU Kernel Programming Language. In \emph{Proceedings of the 7th workshop of Declarative aspects and applications of multicore programming}, DAMP '12. ACM. 
\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran. Efficient Counting Sort Implementations using an Embedded GPU Programming Language. \emph{2nd Workshop of Functional High-Performance Computing, FHPC '13.}
\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
\item Bo Joel Svensson and Ryan Newton. Programming Future Parallel Architectures with Haskell and ArBB. \emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
\item Bo Joel Svensson and Mary Sheeran. Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks. \emph{Proceedings of the 1st ACM SIGPLAN Workshop of Functional High-Performance computing, FHPC '12.}
\end{enumerate}

@@

\subsubsection{Papers B, C, D and E} 
The Obsidian language has grown out of discussions between Mary Sheeran, Koen Claessen and me. 
When it comes to implementation of the ideas we had in these discussions, I have been responsible. 
Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
implementation of them in the setting of Obsidian was done by me. Paper D also have large 
contributions by Mary Sheeran when it comes to examples and implementing them.

Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
variation of counting sort and wanted to implement it. Together we added atomic operations to 
Obsidian and I modified the code generator to accomodate them. Sections 
1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
the benchmarking. Section 8 is joint work.
%Me and Josef are equal partners in this paper and got much valuable advice from 
%Mary Sheeran.

\subsubsection{Papers G and H} 
Here, again, most implementation effort was done by me. For paper G this work was performed 
under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
paper. Programming and running time experimentation was performed by me while getting much 
valuable advice and guidance from Ryan.

The work behind paper H was done by me as a hobby project and my aspirations were not 
high. I just wanted to experiment with some embedded languages techniques that we had 
been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
the right amount of pressure leading to us encompassing much more of ArBB functionality than 
I had in mind. The paper's text is joint work; most examples as well as benchmarking is my work
but the sparse matrix vector multiplication example is entirely Mary's.

%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a fuctioning 
%code generating backend u

@@

This thesis is about applying embedded language techniques to programming 
multi- and many-core computers. Computers today are becoming more and more  
parallel. With each new generation of 
general purpose processor (CPU) or graphics processor (GPU) we get more 
parallel resources. To fully make use of modern CPUs and GPUs we need to
exploit their parallelism.

While it is sometimes possible to automatically parallelise programs originally 
written for serial execution, it is becoming clear that explicitly parallel 
programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
This also means that it is becoming important to think in parallel from the beginning.

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
of cores and number of vector units, varies over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
can use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approach this problem by providing a high level programming interface~\citet{ARBB2011}.
This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++ but a low-level C interface to 
its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or accelerator (Such as the Intel Larrabee). In 
chapter~\ref{chap:ArBB}, there are papers describing work towards embedding Intel 
ArBB in Haskell.

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area for caches, branch prediction and 
ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application.

@@

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
The programmer needs to implement them using lower level features making CUDA and OpenCL 
code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
naive reduction kernel is implement and it is refined over many steps, taking many 
characteristics of the GPU into mind. The end result is a reduction kernel that performs 
very well on the chosen GPU. But as GPUs evolve the design space may need to be explored again. 
%What is 
%is that says that we do not need to redo this exploration with each new GPU generation 
%and discover a new optimal solution?

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still make building blocks implemented by the programmer reusable and composable. 
We want to give more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

@@

\section{Embedded languages}

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Being embedded allows 
us more rapid prototyping of ideas. Other benefits of being embedded often listed 
are: making use of the host language's type system, parser and libraries (to some extent) 
and to have the host language as a power macro language for your embedded language.

A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embedd a compiler 
backend and generate efficient code in some target language. One key aspect of this 
approach is that the embedded language is a library of functions that creates and 
composes Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
embedding that builds ASTs is called a deep embedding. Shallow embeddings on the 
other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
Josef Svenningsson and Emil Axelsson combines the two methods, shallow and deep, 
in order to simplify the AST data type (less constructors) and make more extensible 
embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings while EmbArBB is a more traditional deep embedding.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, there is Lava and Wired in the fields of hardware design and 
verification and Feldspar for digital signal processing\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
programming there are numerous embedded approaches using various host languages. 
But to mention two that use Haskell we have Accelerate and Nikola 
\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
goal of being parallel across platforms \citet{ARBB2011}.

\section{Wider context}
\FloatBarrier

In figure~\ref{fig:researchmatrix} I place our work in relation to other languages,
embedded languages and libraries for parallel programming. The systems are roughly divided 
into three groups based on their level of abstraction. There are low-level languages, here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that go a bit further, they have higher level abstractions and are more easily composable. 
In the highest level I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA the programmer has full controll of how to divide the computation amongst threads and 
blocks. While with Accelerate, Nikola and Thrust the programmer would use high-level patterns
without direct insight into how the application would be mapped onto threads and blocks of the 
GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
but give the programmer more control of what the computation will look like on the GPU. This 
is done by allowing the programmer to specify and compose thread-level and block-level code. 
Recently the CUB library arrived that has a similar idea, that GPU programmers need to 
have block/thread level control to get maximum performance, while maintaining a higher level 
and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
we were alone at targetting this level.

The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
compose to build his application. I place the ArBB/EmbArBB box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate have a single higher order reduction 
operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. I will go into 
what benefits there are to having ArBB embedded in Haskell compared to C++ in 
section~\ref{sec:EmbArBB}.

@@

\FloatBarrier
\section{Obsidian: An embedded language for GPU kernel implementation}

Obsidian is aspiring to raise the level of abstraction of GPU kernel implementation 
while maintaining enough control over details that influence performance on a GPU. 
We try to reach these goals by offering abstractions to the programmer that are 
compositional and parallel. We also ensure that these can easily be compiled 
to NVIDIA CUDA code. We have two different array representations, each with 
its strengths and weaknesses. We also feel that it is important that the programmer 
is able to use local shared memory in computations and has the ability to influence 
memory access patterns. What these array representations and abstractions are 
will be explained in the following sections with references to the papers 
that introduce them or explains them the best.

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy leading to warp programming being a special case.

{\tt ForAll} takes a number of parallel iterations, a body representation by higher order 
syntax. The body is a {\tt Program} at a some level {\tt t} while the resulting program 
is at a level step above. This means a thread program can be turned into a block program 
by using{\tt ForAll} or that a block program can be turned into a grid program.

\subsection{Scalar language}

\subsection{Array representation}

\subsubsection{Pull arrays}
\emph{fusion for free}

\subsubsection{Push arrays}
\emph{fusion for free}

\subsection{Compositionality}

\subsection{Examples} 
\noindent\emph{Example program, use something from latest paper}

\section{Embedding Intel Array Building Blocks}

Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
ability to generate code for a variety of different hardware confgurations. It is implemented 
as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
that while high-performance computing specialist are highly competent at implementing 
kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
a more composable way to write programs that make use of core and vector (SIMD) parallelism and 
doing this while using a familiar language, C++.

ArBB is based on a set of parallel primitives (or structures) on dense 1,2 and 3 dimensional 
arrays and nested arrays: 
\begin{itemize} 
\item {\bf Reductions}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting}      {\tt sort} 
\item {\bf Permutations} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential controlflow} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil}  {\tt map} 
\end{itemize}

>>>>

\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Programming}
\newcommand{\dept}{Department of Computer Science and Engineering}
\newcommand{\uni}{Chalmers University of Technology and G\"oteborg University}
\newcommand{\group}{Functional Programming Research Group}

@@

\begin{enumerate}[A.] 
\item Josef Svenningsson and Bo Joel Svensson. Simple and Compositional Reification of Monadic Embedded Languages, 2013. \emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
\item Bo Joel Svensson, Mary Sheeran and Koen Claessen. Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors. In \emph{Proc. of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
\item Bo Joel Svensson, Koen Claessen and Mary Sheeran. GPGPU kernel implementation and refinement using Obsidian. \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
\item Koen Claessen, Mary Sheeran and Bo Joel Svensson. Expressive Array Constructs in an Embedded GPU Kernel Programming Language. In \emph{Proceedings of the 7th workshop of Declarative aspects and applications of multicore programming}, DAMP '12. ACM. 
\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran. Efficient Counting Sort Implementations using an Embedded GPU Programming Language. \emph{2nd Workshop of Functional High-Performance Computing, FHPC '13.}
\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
\item Bo Joel Svensson and Ryan Newton. Programming Future Parallel Architectures with Haskell and ArBB. \emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
\item Bo Joel Svensson and Mary Sheeran. Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks. \emph{Proceedings of the 1st ACM SIGPLAN Workshop of Functional High-Performance computing, FHPC '12.}
\end{enumerate}

@@

\subsubsection{Papers B, C, D and E} 
The Obsidian language has grown out of discussions between Mary Sheeran, Koen Claessen and me. 
When it comes to implementation of the ideas we had in these discussions, I have been responsible. 
Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
implementation of them in the setting of Obsidian was done by me. Paper D also has large 
contributions by Mary Sheeran when it comes to examples and implementing them.

Paper E was the first I wrote with Josef Svenningsson, who had been thinking about an interesting
variation of counting sort and wanted to implement it. Together we added atomic operations to 
Obsidian and I modified the code generator to accommodate them. Sections 
1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
the benchmarking. Section 8 is joint work.
%Me and Josef are equal partners in this paper and got much valuable advice from 
%Mary Sheeran.

\subsubsection{Papers G and H} 
Here, again, most implementation effort was done by me. For paper G, this work was performed 
under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
paper. Programming and running time experimentation was performed by me, while getting much 
valuable advice and guidance from Ryan.

The work behind paper H was done by me as a hobby project and my aspirations were not 
high. I just wanted to experiment with some embedded languages techniques that we had 
been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
but the sparse matrix vector multiplication example is entirely Mary's.

%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a functioning 
%code generating backend u

@@

This thesis is about applying embedded language techniques to programming 
multi- and many-core computers. Computers today are becoming more and more  
parallel. With each new generation of 
general purpose processor (CPU) or graphics processor (GPU), we get more 
parallel resources. To fully make use of modern CPUs and GPUs, we need to
exploit their parallelism.

While it is sometimes possible to automatically parallelise programs originally 
written for serial execution, it is becoming clear that explicitly parallel 
programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
This also means that it is important to think in parallel from the beginning.

In this thesis the focus is on data-parallelism, as opposed to task- or 
control-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. \emph{more, and maybe not here}

Modern CPUs are parallel in a number of different ways. They discover instruction
level parallelism (ILP), they have multiple processor cores and vector units for 
Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
of cores and number of vector units, varies over available processor models. 
To obtain maximum performance, with a specific target processor in mind, the programmer
can use low-level intrinsics in order to get SIMD parallelism and a threading library
for parallelism over the available cores. One problem with this is that the resulting program 
would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
This interface is based on a set of parallel operations on arrays, such as map, folds 
and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
towards embedding Intel ArBB in Haskell.

Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
For example, where CPUs dedicate large portions of chip area to caches, branch prediction and 
ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
perspective and memory access patterns can make or break the performance of an 
application.

@@

The patterns are powerful abstractions, but it is also important not to be locked into 
a single implementation of each pattern. This could have negative performance impacts. 
In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
The programmer needs to implement them using lower level features, making CUDA and OpenCL 
code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
naive reduction pattern is implemented and then refined over many steps, taking many 
characteristics of the GPU into mind. The end result is an implementation of the 
reduction pattern that performs very well on the chosen GPU. But as GPUs evolve, the 
design space may need to be explored again. 
%What is 
%is that says that we do not need to redo this exploration with each new GPU generation 
%and discover a new optimal solution?

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to give more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Embedded languages
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Embedded languages}

This thesis applies embedded language methodology to data-parallel programming. An 
embedded language is implemented as a library in a host language, Haskell in this case. 
The embedded language approach is beneficial since we are not entirely sure how 
to program current and future highly parallel computers. Being embedded allows 
us more rapid prototyping of ideas. Other benefits of being embedded often listed 
are: making use of the host language's type system, parser and libraries (to some extent) 
and to have the host language as a powerful macro language for your embedded language.

A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embedd a compiler 
backend and generate efficient code in some target language. One key aspect of this 
approach is that the embedded language is a library of functions that create and 
compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
embedding that builds ASTs is called a deep embedding. Shallow embeddings on the 
other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
in order to simplify the AST data type (fewer constructors) and make more extensible 
embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
and deep embeddings, while EmbArBB is a more traditional deep embedding.

The embedded language approach has become popular to use as a way of raising the 
level of abstraction in fields where the default is to use low-level languages. 
For example, there is Lava and Wired in the fields of hardware design and 
verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
programming there are numerous embedded approaches using various host languages. 
But to mention two that use Haskell we have Accelerate and Nikola 
\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
goal of being parallel across platforms \citet{ARBB2011}.
Related work and information about the tools, libraries and language that make the 
wider context surrounding Obsidian is described in section~\ref{sec:relatedwork}.

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Obsidian
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Obsidian: An embedded language for GPU kernel implementation}

Obsidian aspires to raise the level of abstraction of GPU kernel implementation 
while maintaining enough control over details that influence performance on a GPU. 
We try to reach these goals by offering abstractions to the programmer that are 
compositional and parallel. We also ensure that these can easily be compiled 
to NVIDIA CUDA code. We have two different array representations, each with 
its strengths and weaknesses. We also feel that it is important that the programmer 
be able to use local shared memory in computations and has the ability to influence 
memory access patterns. What these array representations and abstractions are 
will be explained in the following sections, with references to the papers 
that introduce them or explain them the best.

The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
that threads can communicate via shared memory without using synchronisation primitives. There 
is no programmer support for warp level programming in CUDA (as there is for block level 
programs). The programmer needs to take care to ensure that the communication patterns are 
within a warp and is then free to remove synchronisations. If warps where to be put in the 
program level hierarchy of Obsidian, its place would be between the thread and block level. 
This forces the programmer to go via warps when putting together a block computation even 
if the communication pattern is not suited for that, leading to inconvenience with no gain.

{\tt ForAll} takes a number of parallel iterations, a body represented by a function 
from an index to a program. The body is a {\tt Program} at a some level {\tt t} while the 
resulting program is at a level step above. This means a thread program can be turned into 
a block program by using {\tt ForAll} or that a block program can be turned into a grid program.

\subsection{Scalars}

Scalars and operations on scalars are represented by an expression data type ({\tt Exp a}) in 
Obsidian. This is another example of a deep embedding; the expression data type contains 
constructors for literal values, arithmetic operations, and conditionals.

Haskell's type class system allows us to overload arithmetic operations such as {\tt (+)}, 
{\tt (-)} and {\tt (*)} by making expressions an instance of {\tt Num}. This allows the 
Obsidian arithmetic expressions to look exactly like corresponding native Haskell arithmetic.

The approach taken to embedd the scalar language is very similar to what is described 
in~\citet{COMPILEEDSL}; but we use a GADT to obtained typed expressions rather than phantom types.

\subsection{Arrays}

Obsidian has two array representations. These array representations are implemented as 
shallow embeddings on top of the expression and program data types. This means that these 
arrays will disappear during Haskell evaluation of the Obsidian program.

\subsubsection{Pull arrays}

Pull arrays have been part of Obsidian from the very beginning~\citet{JMT} but called either 
{\tt Array} or just {\tt Arr}. A pull array represents an array as an indexing function; given 
an index it provides an element:

\begin{verbatim} 
data Pull s a = Pull {pullLen :: s, 
                      pullFun :: EWord32 -> a}
\end{verbatim}

Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
or dynamic (length represented by an expression, an unknown length). When creating block and 
warp level computations we want to know the exact lengths in order to avoid generating code 
riddled with conditionals concerning array lengths. Conditionals are problematic on GPUs and 
while general programs that run for any size is possible when performance is crucial a fixed 
size is recommended.

One benefit of pull arrays are that they give fusion of all operations on them for free. for example {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:

\begin{verbatim} 
map   f (Pull n ixf) = Pull n (f . ixf)
\end{verbatim}

Now, we see that {\tt map f . map g} is the same as {\tt map f . g} since both result in {\tt f}
and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
memory.

Another benefit of pull arrays are that they are parallel. Simply put, early Obsidian was pull 
arrays in addition to a way to compute the array and store its elements to memory 
(called {\tt sync} in early versions of Obsidian and later {\tt force}). A very 
direct way to compile a pull array to a GPU is to launch as many threads as there are 
elements in the array and apply the indexing function to the thread id. Now each thread 
will have computed an element of the array and can store that to memory.

\subsubsection{Push arrays}

Push arrays where added to Obsidian as a complement to pull arrays. Pull arrays are 
good because they are so simple to understand and easy to parallelise. However certain
operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
or interleaving of pull arrays result in conditionals in the indexing function. The 
interleaving case, which is worst, leads to diverging branching in every warp. If branches 
diverge the GPU will execute turn off the processing elements taking one path and 
allow the others to progress; then turning to the other branch. The computation of the 
branches has become serialised and compute resources are wasted.

\begin{verbatim}
data Push p s a =
  Push s ((a -> EWord32 -> TProgram ()) -> Program p ())
\end{verbatim}

\emph{point to results in each paper}

\emph{provide information around the results in papers}

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% ArBB work
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Embedding Intel Array Building Blocks}

Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
ability to generate code for a variety of different hardware configurations. It is implemented 
as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
that while high-performance computing specialists are highly competent at implementing 
kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
a more composable way to write programs that make use of core and vector (SIMD) parallelism and 
doing this while using a familiar language, C++.

ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
\begin{itemize} 
\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
\item {\bf Sorting:}      {\tt sort} 
\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
\item {\bf Map, zipWith, stencil:}  {\tt map} 
\end{itemize}

These operations (and many more, there are about 120 operations in total) can be used 
by the programmer when writing programs using ArBB. Note that there are sequential looping 
constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
language and calling the ArBB functions does not actually execute them, it just ads a node to 
some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
This is exactly the way these things work in Haskell embedded languages, but it may be even 
more surprising to a C++ programmer. Before an ArBB function can be computed it needs to be 
captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
functionality. The first time {\tt call} is used on a function pointer that function is 
executed and the ArBB functions used in it build an AST (we say the function has been 
captured). The AST is then compiled and specialised for the hardware available. Any 
subsequent calls of a function using ArBB functionality do not lead to a recompilation, 
the first call caches this compiled code.

The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
is not useful for application implementation. A higher-level interface to ArBB to offer 
the Haskell programmer is needed. Our first attempt was to implement an Accelerate backend 
using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
more richer language and there was an API mismatch between Accelerate and ArBB. These problems 
are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
to the Haskell in a way that is useful and simpler to implement, we start working on EmbArBB.

\subsubsection{Operations on arrays}

EmbArBB is a traditional deep embedding, there is an AST data type with constructors 
representing each ArBB language feature. It is true that a deep embedding of ArBB 
leads to duplication of effort. The low-level operations exported by the 
ArBB C interface already construct an AST within the ArBB system. One possible benefit 
of the deep embedding is that it buys independence from the ArBB backend. It is possible 
to compile the AST generated on the Haskell side to some other target language.

When compiling EmbArBB to ArBB a sharing detection pass is performed. This pass detects 
sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
is that it reduces the numbers of calls into the ArBB C library. The efficiency of code 
generated is probably not affected since ArBB would discover that sharing in a Common 
Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
on the Haskell side results in a simpler AST being built on the ArBB side; potentially making 
the job for the ArBB optimisation passes simpler. What effects these 
operations truly had on performance has not been investigated. Actually, applying a sharing 
detection technique was one of my personal motivations for working on EmbArBB. In Obsidian 
we have not seen the need for this technique.

One benefit of a deep embedding is that transformations can be performed on the AST. Applying 
any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
Therefore we apply no further transformations after sharing detection. However, if EmbArBB 
would be used to generate code in some other target language a set of optimisations 
would be needed.

In paper H ,section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
the code, that code is identical no matter if the AST was build via Haskell or C++ calls. 
Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
takes to go through the process of making those ArBB C API calls. Although if the generated 
code is used many times the caching ArBB does of generated code means that the initial code 
generation cost can be amortised.

Since our work with ArBB, Intel has unfortunately retired the ArBB system. This 
leaves EmbArBB without a functioning code generating backend. This makes the choice of a
deep embedding (in hindsight) the correct one. EmbArBB could be resurrected by implementing 
a new code generating backend.

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Related work
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\section{Related work} 
\label{sec:relatedwork}
\FloatBarrier

In figure~\ref{fig:researchmatrix} I our work is placed in relation to other languages,
embedded languages and libraries for parallel programming. The systems are roughly divided 
into three groups, based on their level of abstraction. There are low-level languages, here 
I place languages that are imperative and C-like. In the middle layer I place languages 
that go a bit further, they have higher level abstractions and are more easily composable. 
In the highest level I place languages that completely abstract away from details of the 
hardware they run on. The division is not free from personal bias and the placement of languages
in boxes was not always easy.

@@

With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
CUDA, the programmer has full control of how to divide the computation amongst threads and 
blocks. Using Accelerate, Nikola and Thrust the programmer would use high-level patterns
without direct insight into how the application would be mapped onto threads and blocks of the 
GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
but give the programmer more control of what the computation will look like on the GPU. This 
is done by allowing the programmer to specify and compose thread-level and block-level code. 
Recently, the CUB library arrived that is built on a similar idea, that GPU programmers need to 
have block/thread level control to get maximum performance, while maintaining a higher level 
and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
we were alone at targeting this level.

The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
they are not general primitives. Where Accelerate have a single higher order reduction 
operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
%I will go into 
%what benefits there are to having ArBB embedded in Haskell compared to C++ in

@@
